---
title: "Script capítulo Temporal"
author: "Ruben"
date: "2025-07-11"
output: html_document
---


## Loading of the libraries and data 

Here we load the libraries we will use in the study. The data generated from sequence processing following the standard workflow of the DADA2 package are loaded. These data contain the table of ASV abundances per sample, as well as their taxonomic assignment. 


```{r}
knitr::opts_knit$set(root.dir = getwd())
suppressMessages(suppressWarnings({
datos <- "dada2multithreads.RData"
suppressPackageStartupMessages({
library(dada2)
library(ggplot2)
library(here)
library(phyloseq)
library(microbiome)
library(dplyr)
library(writexl)
library(readxl)
library(openxlsx)
library(gridExtra)
library(patchwork)
library(Biostrings)
library(tidyr)
library(tibble)
library(vegan)
library(scales)
library(car)
library(officer)
library(flextable)
library(patchwork)
library(broom)
library(pheatmap)
library(picante)
library(reshape2)
library(ppcor)
library(lmtest)
library(sandwich)
library(forcats)
library(Maaslin2)
library(lme4)
})
theme_set(theme_bw())
load(here("data",datos)) })) 

dir.create("test")
dir.create("graficos")
out_graf <- "graficos"
if (!dir.exists(out_graf)) dir.create(out_graf)

```

## Data preparation
  
1. All experimental data were imported from an Excel file.
2. A phyloseq object was created.

```{r}
samdf <- read_excel("metadata_Temp.xlsx")
samdf <- as.data.frame(samdf)


rownames(samdf) <- samdf$Nombre

ps <- phyloseq(otu_table(seqtab_nochim, taxa_are_rows = FALSE), 
               sample_data(samdf), 
               tax_table(tax))

```

### Removal of controls and non-representative samples

Samples corresponding to controls were removed from the dataset, including negative plant controls (CN), DNA extraction controls (Cext), and negative PCR controls (NegPCR). Samples obtained from plants that did not grow properly were also excluded.

```{r}

ps_nocontrol <- subset_samples(ps, !(Comunidad %in% c("CN", "Cext", "NegPCR"))& 
                 Observaciones %in% c("HONGO", "OK"))

```

### Removal of non-bacterial sequences

Sequences corresponding to mitochondria, chloroplasts, and eukaryotes were removed. This was done because the selected primers may also amplify the 16S region of these organelles.

```{r}
print("Initial:")
ps_nocontrol
inicial<-sum(otu_table(ps_nocontrol))
print ("Here, chloroplasts, mitochondria, and eukaryotic sequences were removed")
ps_nocontrol_filtrado<- subset_taxa(ps_nocontrol, Class!="Chloroplast")
sin_cloroplastos<-sum(otu_table(ps_nocontrol_filtrado))
ps_nocontrol_filtrado <- subset_taxa(ps_nocontrol_filtrado, Family!="Mitochondria")
sin_mitocondrias<-sum(otu_table(ps_nocontrol_filtrado))
ps_nocontrol_filtrado <- subset_taxa(ps_nocontrol_filtrado, Kingdom!="Eukaryota")
sin_eucariotas<-sum(otu_table(ps_nocontrol_filtrado))
valores<-c(inicial,sin_cloroplastos,sin_mitocondrias,sin_eucariotas)
print("initial, after removal of chloroplasts, after removal of mitochondria, after removal of eukaryotes")
print(valores)
print(paste0("Total Initial= ",inicial))
print(paste0("  removing chloroplasts= ",sin_cloroplastos))
print(paste0("  removing chloroplasts and mitochondria= ",sin_mitocondrias))
print(paste0("  removing chloroplasts, mitochondria, and eukaryotes= ",sin_eucariotas))
print("Filtered:")
ps_nocontrol_filtrado

suppressMessages(summarize_phyloseq(ps_nocontrol_filtrado))
```


### Removal of outliers

Previous analyses showed that some samples from community E had a microbial composition that was markedly different from the rest of the samples in that community. The 10 most abundant families in these samples were calculated and compared with those in other samples to confirm this difference. The reasons why these samples failed are unknown, but given that their profiles were very different from the rest, they were excluded from the study to avoid biasing the results.


```{r}

# Selection of community E only
ps_E <- subset_samples(ps_nocontrol_filtrado, Comunidad == "E")

# Family-level aggregation
ps_familia_E <- tax_glom(ps_E, taxrank = "Family")

# Convert to relative abundances
ps_familia_E_rel <- transform_sample_counts(ps_familia_E, function(x) x / sum(x))

# Convert to data frame
df_familia_E <- as.data.frame(psmelt(ps_familia_E_rel))
df_familia_E$Family <- as.character(df_familia_E$Family)

# Sum abundances by family
df_familia_suma_E <- df_familia_E %>%
  group_by(Family) %>%
  summarise(Total = sum(Abundance), .groups = "drop") %>%
  arrange(desc(Total))

# Select top 10 families
top10_familias_E <- df_familia_suma_E$Family[1:10]

# Filter and prepare for plotting
df_top10_E <- df_familia_E %>%
  filter(Family %in% top10_familias_E) %>%
  group_by(Sample, Family) %>%
  summarise(Abundancia = sum(Abundance) * 100, .groups = "drop")

# Plot
Plot_top10_E <- ggplot(df_top10_E, aes(x = Sample, y = Abundancia, fill = Family)) +
  geom_bar(stat = "identity") +
  theme_bw() +  
  ylab("Abundancia relativa (%)") +
  xlab("Muestras") +
  ggtitle("Top 10 familias - Comunidad E") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("graficos/top10_familias_ComunidadE.png",
       plot = Plot_top10_E,
       width = 8, height = 6, dpi = 300)

outliers <- c("Temp_E_7_1_S219", "Temp_E_5_1_S146", "Temp_E_8_1_S4")

ps_nocontrol_filtrado_2 <- subset_samples(ps_nocontrol_filtrado, !(sample_names(ps_nocontrol_filtrado) %in% outliers))

```

### Correction of NAs 

Samples with MPN values recorded as NA, either because the corresponding bacteria did not grow in those samples or because the values were below the detection limit (<110000 MPN/mL), were converted to the detection-limit value of 220000 MPN per plant.

1. The text values ‘NA’ are replaced with missing values, which are coded as NA, but not in text format as they appear in the original data sheet. 
2. The MPN column was converted to numeric type.
The NA (missing) values in the MPN column are replaced with 220000.
4. All values below 110000 in the MPN column were converted to 220000.

For dry weight, samples with NA values (because the aboveground part did not grow sufficiently) were assigned a value of 0 g.

1. The text values ‘NA’ are replaced with NA (missing) values.
2. The Weight column was converted to numeric type.
3.  The NA (missing) values in the Weight column are replaced with 0.

```{r}

sample_data(ps_nocontrol_filtrado_2)$MPN[sample_data(ps_nocontrol_filtrado_2)$MPN == "NA"] <- NA
sample_data(ps_nocontrol_filtrado_2)$MPN <- as.numeric(sample_data(ps_nocontrol_filtrado_2)$MPN)
sample_data(ps_nocontrol_filtrado_2)$MPN[is.na(sample_data(ps_nocontrol_filtrado_2)$MPN)] <- 220000
sample_data(ps_nocontrol_filtrado_2)$MPN[sample_data(ps_nocontrol_filtrado_2)$MPN < 220000] <- 220000

sample_data(ps_nocontrol_filtrado_2)$Peso_Seco[sample_data(ps_nocontrol_filtrado_2)$Peso_Seco == "NA"] <- NA
sample_data(ps_nocontrol_filtrado_2)$Peso_Seco <- as.numeric(sample_data(ps_nocontrol_filtrado_2)$Peso_Seco)
sample_data(ps_nocontrol_filtrado_2)$Peso_Seco[is.na(sample_data(ps_nocontrol_filtrado_2)$Peso_Seco)] <- 0



sample_data(ps_nocontrol_filtrado_2)$MPN.Peso[sample_data(ps_nocontrol_filtrado_2)$MPN.Peso == "NA"] <- NA
sample_data(ps_nocontrol_filtrado_2)$MPN.Peso <- as.numeric(sample_data(ps_nocontrol_filtrado_2)$MPN.Peso)
sample_data(ps_nocontrol_filtrado_2)$MPN.Peso[is.na(sample_data(ps_nocontrol_filtrado_2)$MPN.Peso)] <- 0

```

## Subsampling 

The number of sequences per sample was visualized to determine the subsampling depth. A fixed seed of 134 was used.

```{r}

sort(sample_sums(ps_nocontrol_filtrado_2), decreasing = TRUE)

suppressMessages(suppressWarnings({
  set.seed(134)  
  ps_nocontrol_filtrado_rare <- rarefy_even_depth(ps_nocontrol_filtrado_2, sample.size = 24562, trimOTUs = TRUE)
}))

png("graficos/profundidad_secuenciacion.png", width = 800, height = 600)

plot(
  sort(sample_sums(ps_nocontrol_filtrado_2)),
  ylab = "Sequencing depth",
  xlab = "Samples ordered by depth",
  main = "Sequencing depth per sample",
  cex.lab = 1.6,   
  cex.axis = 1.4,   
  cex.main = 1.8    
)

abline(h = 24562, col = "red", lwd = 2)

dev.off()

```

## Saving and loading 

Saving of the phyloseq object and loading of data with phylogenetic distance.


```{r}

saveRDS(ps_nocontrol_filtrado_rare, "ps_nocontrol_filtrado_rare.rds")
# In a separate script, the sequences from the previously generated phyloseq object were matched against a reference database with an associated phylogenetic tree and taxonomy (e.g., Greengenes2). This allowed the phylogenetic tree to be incorporated into the object and enabled analyses based on phylogenetic metrics (e.g., Faith's PD, UniFrac, etc.).
nuevosdatos <- "phyloseq_objects_with_tree.Rdata"

load(here("data",nuevosdatos ))

```

---

# **Diversity analysis and correlations** 

## Creation of working variables 

```{r}

datos_est <- as(sample_data(ps_nocontrol_filtrado_rare), "data.frame")
datos_est$log_MPN <- log10(datos_est$MPN)
datos_est$log_MPN.Peso <- datos_est$log_MPN / datos_est$Peso_Seco

Riqueza <- estimate_richness(ps_nocontrol_filtrado_rare, measures = "Observed")
Homogeneidad <- estimate_richness(ps_nocontrol_filtrado_rare, measures = "Shannon")

Faith_D <- pd(as(otu_table(new_ps), "matrix"), phy_tree(new_ps))
datos_est$Riqueza <- Riqueza[rownames(datos_est), "Observed"]
datos_est$Homogeneidad <- Homogeneidad[rownames(datos_est), "Shannon"]
datos_est$Faith_D <- Faith_D[rownames(datos_est), "PD"]

# Variable type checking

datos_est$Comunidad   <- factor(datos_est$Comunidad)   # unordered factor
datos_est$Tiempo  <- as.integer(datos_est$Tiempo)  # 1..8
# Centered and scaled version of time for numerical stability
datos_est$cTiempo <- as.numeric(scale(datos_est$Tiempo, center = TRUE, scale = TRUE))

# Sum-to-zero contrast coding is recommended for stable interpretation in unbalanced designs. Factors are coded so that the intercept represents the overall mean across levels, and coefficients represent deviations from that mean rather than comparisons with a reference level. This yields more stable and interpretable results in unbalanced designs.
options(contrasts = c("contr.sum", "contr.poly"))

```

### Visualization of working variables

The values of the working variables are plotted by Community and Time.

```{r}

# Means ± error by Time and Community

vars <- c("log_MPN", "Peso_Seco", "Riqueza", "Homogeneidad", "Faith_D")

lista_graficos <- list()

for (v in vars) {
  # Summary by Community × Time
  df_mean <- datos_est %>%
    group_by(Comunidad, Tiempo) %>%
    summarise(
      n    = n(),
      mean = mean(.data[[v]], na.rm = TRUE),
      sd   = sd(.data[[v]],   na.rm = TRUE),
      se   = sd / sqrt(n),
      .groups = "drop"
    )

# Plot: means ± SE by Time and Community
  p <- ggplot(df_mean, aes(x = Tiempo, y = mean, color = Comunidad, group = Comunidad)) +
    geom_line(size = 1) +
    geom_point(size = 2) +
    geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.15) +
    labs(
      x = "Time",
      color = "Community",
      y = ifelse(v == "Riqueza", "Observed richness",
      ifelse(v == "Peso_Seco", "Dry Weight (g)",
      ifelse(v == "Homogeneidad", "Evenness",
      ifelse(v == "log_MPN.Peso", "log_MPN/Peso",
      ifelse(v == "Faith_D", "Faith’s diversity", v))))) 
  ) +
    theme_bw()+
      scale_x_continuous(breaks = unique(df_mean$Tiempo))+
  theme(
    axis.text.x = element_text(size = 10, angle = 0, hjust = 1),  
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 10, face = "bold"),         
    axis.title.y = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12, face="bold"),
    plot.title   = element_text(size = 12, face = "bold", hjust = 0.5) 
  )


    lista_graficos[[v]] <- p

  fn <- file.path(out_graf, paste0( "medias_SE_", v, ".png"))
  ggsave(filename = fn, plot = p, width = 7, height = 5, dpi = 300)

  message("Guardado: ", fn)
}

panel <- (
  lista_graficos[["log_MPN"]] + labs(title = NULL) +
  lista_graficos[["Peso_Seco"]] + labs(title = NULL)
) +
  plot_annotation(tag_levels = "A", tag_suffix = ") ") & 
  theme(legend.position = "right")

panel <- panel + plot_layout(guides = "collect")


ggsave("graficos/medias_SE_logMPN_PesoSeco.png",
       plot = panel,
       width = 10, height = 5, dpi = 300)


panel <- (
  lista_graficos[["Riqueza"]] + labs(title = NULL) +
  lista_graficos[["Homogeneidad"]] + labs(title = NULL)
) +
  plot_annotation(tag_levels = "A", tag_suffix = ") ") & 
  theme(legend.position = "right")

panel <- panel + plot_layout(guides = "collect")


ggsave("graficos/medias_SE_Riqueza_Homogeneidad.png",
       plot = panel,
       width = 10, height = 5, dpi = 300)

```

The ratio between load and weight values is represented here. Data with a weight of 0 are removed, since the ratio cannot be calculated for them.

```{r}


datos_est2 <- subset(datos_est, is.finite(log_MPN.Peso))

df_mean_log_MPN.Peso <- datos_est2 %>%
  group_by(Comunidad, Tiempo) %>%
  summarise(
    n    = n(),
    mean = mean(.data[["log_MPN.Peso"]], na.rm = TRUE),
    sd   = sd(.data[["log_MPN.Peso"]], na.rm = TRUE),
    se   = sd / sqrt(n),
    .groups = "drop"
  )

medias_log_MPN.Peso <- ggplot(df_mean_log_MPN.Peso, aes(x = Tiempo, y = mean, color = Comunidad, group = Comunidad)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.15) +
  labs(
    x = "Time",
    y = "log10(MPN) / Peso seco"
  ) +
  theme_bw() +
  scale_x_continuous(breaks = unique(df_mean_log_MPN.Peso$Tiempo)) +
  theme(
    axis.text.x = element_text(size = 10, angle = 0, hjust = 1),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 10, face = "bold"),
    axis.title.y = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )

# guardar archivo
ggsave("graficos/medias_log_MPN.Peso.png", plot = medias_log_MPN.Peso,
       width = 7, height = 5, dpi = 300)

message("Guardado: ", fn)
```


### Statistical analysis of Time and Community 

The statistical effects of Time, Community, and their interaction on the working variables were analyzed.

```{r}

vars <- c("log_MPN", "Peso_Seco", "Riqueza", "Homogeneidad", "Faith_D")
mods <- list()

datos_est$Comunidad <- relevel(datos_est$Comunidad, ref = "A")
contrasts(datos_est$Comunidad)
# Community 1 corresponds to Community A, Community 2 to Community B, and so on.

pdf("Resultados_Modelos_Graficos.pdf", width = 12, height = 8)
for (y in vars) {
  cat("\n\n====================================================\n")
  cat("Respuesta:", y, "\n")
  form <- as.formula(paste0(y, " ~ cTiempo * Comunidad"))
  fit  <- lm(form, data = datos_est)
  mods[[y]] <- fit
  
  cat("\n--- SUMMARY (individual coefficients vs. reference) ---\n")
  print( lmtest::coeftest(fit, vcov = sandwich::vcovHC(fit, type = "HC3"))
)
  
  cat("\n--- ANOVA global (Type III de base R) ---\n")
  print(Anova(fit,type = 3, white.adjust = "hc3"))
  
  # Diagnostic plots
  par(mfrow=c(2,2))
  plot(fit, main = y)
  par(mfrow=c(1,1))
}


for (v in vars) {
  cat("\n====================================================\n")
  cat("Variable:", v, "\n")
  
  m <- mods[[v]]  # retrieve the model
  residuales <- residuals(m)
  
    # Shapiro-Wilk test for residual normality
  cat("\n--- Shapiro-Wilk (normalidad de residuales) ---\n")
  if (length(residuales) <= 5000) {
    print(shapiro.test(residuales))
  } else {
    cat("Too many observations for Shapiro (>5000), skipping the test.\n")
  }
  
  # Levene's test for homogeneity of variances across Communities
  cat("\n--- Levene's test (residual variances across Communities) ---\n")
  print(leveneTest(residuales ~ datos_est$Comunidad))
  
  # Breusch-Pagan test for heteroscedasticity
  cat("\n--- Breusch-Pagan test (model heteroscedasticity) ---\n")
  print(lmtest::bptest(m))
}

dev.off()

```

A linear model was fitted for the load-to-weight ratio as a function of Community, Time, and their interaction. Data with Weight = 0 were excluded because the ratio could not be calculated.

```{r}

datos_est2 <- subset(datos_est, is.finite(log_MPN.Peso))


datos_est2$Comunidad <- relevel(datos_est2$Comunidad, ref = "A")
contrasts(datos_est2$Comunidad)
# Community 1 corresponds to Community A, Community 2 to Community B, and so on.

pdf("Resultados_log_MPN_Peso.pdf", width = 12, height = 8)

cat("\n\n====================================================\n")
cat("Respuesta: log_MPN.Peso\n")
lm_log_MPN.Peso <- lm(log_MPN.Peso ~ cTiempo * Comunidad, data = datos_est2)

cat("\n--- SUMMARY (individual coefficients vs. reference) ---\n")
print(lmtest::coeftest(lm_log_MPN.Peso, vcov = sandwich::vcovHC(lm_log_MPN.Peso, type = "HC3")))

cat("\n--- ANOVA global (Type III de base R) ---\n")
print(Anova(lm_log_MPN.Peso, type = 3, white.adjust = "hc3"))

# Diagnostic plots 
par(mfrow=c(2,2))
plot(lm_log_MPN.Peso, main = "log_MPN.Peso")
par(mfrow=c(1,1))

residuales <- residuals(lm_log_MPN.Peso)

# Test de normalidad de residuales (Shapiro-Wilk)
cat("\n--- Shapiro-Wilk (normalidad de residuales) ---\n")
if (length(residuales) <= 5000) {
  print(shapiro.test(residuales))
} else {
  cat("Too many observations for Shapiro (>5000), skipping the test.\n")
}

 # Levene's test for homogeneity of variances across Communities
cat("\n--- Levene's test (residual variances across Communities) ---\n")
print(leveneTest(residuales ~ datos_est2$Comunidad))

# Breusch-Pagan test for heteroscedasticity
cat("\n--- Breusch-Pagan test (model heteroscedasticity) ---\n")
print(lmtest::bptest(lm_log_MPN.Peso))

dev.off()


```

### Effect of Richness and phylogenetic diversity on Load and Weight 

The effect of Richness and phylogenetic diversity on Load and Weight was assessed. First, the correlation between both variables was examined, and if one was detected, the fraction of phylogenetic diversity independent of Richness was calculated. Next, the effects of Richness and the fraction of phylogenetic diversity independent of Richness on Load and Weight were evaluated.

```{r}

# 1) Initial exploration ---------------------------------------------------
# Scatterplot of Richness vs Faith_D with smoothing and a fitted line to assess linearity
png("graficos/riqueza_vs_faith.png", width = 800, height = 600)

par(mar = c(5, 6.5, 4, 2) + 0.1,   
    mgp = c(2.5, 0.8, 0))          

plot(datos_est$Riqueza, datos_est$Faith_D,
     xlab = "Richness", ylab = "Diversity",
     main = "Richness vs. Diversity",
     cex.lab = 1.8,   
     cex.axis = 1.6,  
     cex.main = 2.2,  
     cex = 1.2,       
     pch = 20, col = "grey30")

lines(lowess(datos_est$Riqueza, datos_est$Faith_D), col = "blue")
abline(lm(Faith_D ~ Riqueza, data = datos_est), col = "red", lwd = 2)

dev.off()

# Correlation (numerical)
res_cor <- cor.test(datos_est$Riqueza, datos_est$Faith_D, method = "pearson")
print(res_cor)          
res_cor$estimate        
res_cor$p.value          


# 2) Residualize Faith_D with respect to Richness (fraction of Faith diversity independent of Richness) ---- ----
m_FD_vs_r <- lm(Faith_D ~ Riqueza, data = datos_est)
datos_est$FD_resid <- resid(m_FD_vs_r)   

# 3) Models with the fraction of Faith diversity independent of Richness

lm_Peso_resid <- lm(Peso_Seco ~ Riqueza + FD_resid, data = datos_est)
cat("Final model summary (Dry Weight ~ Richness + FD_resid):\n")
print(summary(lm_Peso_resid))
lmtest::bptest(lm_Peso_resid)

lm_MPN_resid <- lm(log_MPN ~ Riqueza + FD_resid, data = datos_est)
cat("Resumen modelo final (log_MPN ~ Riqueza + FD_resid):\n")
print(summary(lm_MPN_resid))
print(Anova(lm_MPN_resid, type = "II"))
lmtest::bptest(lm_MPN_resid)

```

# **Phylogenetic analysis** 

## Top 10 families 

The 10 most abundant families were calculated.

```{r}

# Because the data were rarefied, this was done as follows
ps_familia <- tax_glom(ps_nocontrol_filtrado_rare, taxrank = "Family")
ps_familia_rel <- transform_sample_counts(ps_familia, function(x) x / sum(x))


# Add "Other" to the plot
df <- phyloseq::psmelt(ps_familia_rel) |>
  mutate(
    Family = as.character(Family),
    Family = ifelse(is.na(Family) | Family == "", "Unassigned", Family),
    Family_plot = forcats::fct_lump_n(Family, n = 10, w = Abundance, other_level = "Otros")
  )


# Sum by sample (Top 10) and add "Other"; order the legend
per_sample <- df |>
  group_by(Sample, Comunidad, Tiempo, Family_plot) |>
  summarise(Abund_rel = sum(Abundance), .groups = "drop")

# Summarize for the plot: mean by Community x Time x Family_plot
familia_plot <- per_sample |>
  group_by(Comunidad, Tiempo, Family_plot) |>
  summarise(NormalizedAbundance = mean(Abund_rel), .groups = "drop")

# Reorder
familia_plot <- familia_plot |>
  mutate(
    Family_plot = forcats::fct_reorder(Family_plot, NormalizedAbundance, .fun = mean, .desc = TRUE),
    Family_plot = forcats::fct_relevel(Family_plot, "Otros", after = Inf))


Top10_familias_otros <- ggplot(familia_plot, aes(x = factor(Tiempo), y = NormalizedAbundance, fill = Family_plot)) +
  geom_bar(stat = "identity", position = "stack")  +
  scale_fill_brewer(palette = "Set3") +
    labs(title = "Top10 famylies",
       x = "Time",
       y = "Relative abundance",
       fill = "Family") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    legend.title = element_text(size = 10, face = "bold")) +
  facet_wrap(~ Comunidad, scales = "free_y", ncol = 3) 

ggsave("graficos/Top10_Familias_otros.png",
       plot = Top10_familias_otros, width = 12, height = 6, dpi = 300)

```

## Effect of time on taxa abundance 

In this section, the effect of time on the abundance of the different phyla and families will be evaluated.

```{r}

# ============================================
#  Phylum-level analysis
# ============================================

ps_filo <- tax_glom(ps_nocontrol_filtrado_rare, taxrank = "Phylum", NArm = FALSE)
ps_filo_rel <- transform_sample_counts(ps_filo, function(x) {
  if (sum(x, na.rm = TRUE) > 0) x / sum(x, na.rm = TRUE) else x
})

# Extract OTU table
otu_filo <- as.data.frame(otu_table(ps_filo_rel))

# Transpose if needed

if(taxa_are_rows(ps_filo_rel)) otu_filo <- as.data.frame(t(otu_filo))

# Extract taxonomy
tax_filo <- as.data.frame(tax_table(ps_filo_rel))
colnames(otu_filo) <- tax_filo$Phylum

# Run MaAsLin2 at phylum level
fit_filo <- Maaslin2(
  input_data = otu_filo,
  input_metadata = datos_est,
  output = "maaslin2_Filo_Tiempo",
  fixed_effects = c("Tiempo"),
  random_effects = c("Comunidad"),
  normalization = "NONE",
  transform = "AST",
  min_abundance = 0.0001,
  min_prevalence = 0.1,
  plot_heatmap = FALSE,
  plot_scatter = FALSE
)

# Read results
maaslin_filo_res <- read.delim("maaslin2_Filo_Tiempo/all_results.tsv")
colnames(maaslin_filo_res)[colnames(maaslin_filo_res) == "feature"] <- "Phylum"
maaslin_filo_sig <- maaslin_filo_res[maaslin_filo_res$qval < 0.05, ]
maaslin_filo_sig$Direction <- ifelse(maaslin_filo_sig$coef > 0, "↑ Increases", "↓ Decreases")

maaslin_filo_sig <- maaslin_filo_sig[order(-abs(maaslin_filo_sig$coef)), ]

print(maaslin_filo_sig[, c("Phylum", "coef", "qval", "Direction")])

# ============================================
#   Family-level analysis
# ============================================

ps_familia <- tax_glom(ps_nocontrol_filtrado_rare, taxrank = "Family")

ps_familia_rel <- transform_sample_counts(ps_familia, function(x) {
  if (sum(x, na.rm = TRUE) > 0) x / sum(x, na.rm = TRUE) else x
})


# Extract OTU table
otu_familia <- as.data.frame(otu_table(ps_familia_rel))

# Transpose if needed

if(taxa_are_rows(ps_familia_rel)) otu_familia <- as.data.frame(t(otu_familia))

# Extract taxonomy
tax_familia <- as.data.frame(tax_table(ps_familia_rel))
colnames(otu_familia) <- tax_familia$Family  


# Run MaAsLin2

fit_familia <- Maaslin2(
  input_data = otu_familia,
  input_metadata = datos_est,
  output = "maaslin2_Familia_Tiempo",
  fixed_effects = c("Tiempo"),
  random_effects = c("Comunidad"),
  normalization = "NONE",  
  transform = "AST",       
  min_abundance = 0.0001,
  min_prevalence = 0.1,
  plot_heatmap = FALSE,
  plot_scatter = FALSE
)

# Read results
maaslin_familia_res <- read.delim("maaslin2_Familia_Tiempo/all_results.tsv")
colnames(maaslin_familia_res)[colnames(maaslin_familia_res) == "feature"] <- "Family"


# Filter significant
maaslin_familia_sig <- maaslin_familia_res[maaslin_familia_res$qval < 0.05, ]
maaslin_familia_sig$Direction <- ifelse(maaslin_familia_sig$coef > 0, "↑ Increases", "↓ Decreases")

cat("Significant families (q < 0.05):", nrow(maaslin_familia_sig), "\n\n")
print(maaslin_familia_sig[, c("Family", "coef", "pval", "qval", "Direction")])

# Families that increase
Familias_aumentan <- maaslin_familia_sig[maaslin_familia_sig$coef > 0, ]
Familias_aumentan <- Familias_aumentan[order(-abs(Familias_aumentan$coef)), ]

# Results
print(Familias_aumentan[, c("Family", "coef", "qval", "Direction")])

# Families that decrease

Familias_disminuyen <- maaslin_familia_sig[maaslin_familia_sig$coef < 0, ]
Familias_disminuyen <- Familias_disminuyen[order(-abs(Familias_disminuyen$coef)), ]

# Results
print(Familias_disminuyen[, c("Family", "coef", "qval", "Direction")])


# Create a list with the results
lista_resultados <- list(
  "Filos_significativos" = maaslin_filo_sig,
  "Aumentan" = Familias_aumentan,
  "Disminuyen" = Familias_disminuyen)

# Save to Excel
write_xlsx(lista_resultados, "Resultados_Filo_Familia_Temporales.xlsx")

```
---

# **Composition analysis**  

## Microbial composition - Bray-Curtis

Here, microbial composition is represented using NMDS (Non-metric Multidimensional Scaling) based on Bray-Curtis distances, separating the data by Community type and sampling time, and visually grouping them according to these variables.


```{r}


ps_prop <- transform_sample_counts(ps_nocontrol_filtrado_rare, function(otu) otu/sum(otu))
set.seed(123)
ord_nmds_bray <- ordinate(ps_prop, method="NMDS", distance="bray")

# First, calculate the distance matrix using Bray-Curtis
bray_Comunidad <- plot_ordination(ps_prop, ord_nmds_bray, shape="Tiempo", color="Comunidad", title="") +
  stat_ellipse(aes(color=Comunidad, group=Comunidad), level=0.95, linetype="solid", linewidth=1) +   
  geom_point() + 
  labs (
    shape = "Time",
     color = "Community")+
  scale_shape_manual(values = c(0:9)) + 
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink", "brown", "yellow", "cyan", "black")) +
  guides(color = guide_legend(nrow = 5), shape = guide_legend(nrow = 5)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 0, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(margin = margin(b = 10)))
  
ggsave("graficos/bray_Comunidad.svg", plot = bray_Comunidad, width = 8, height = 6, dpi = 300)

bray_tiempo <- plot_ordination(ps_prop, ord_nmds_bray, shape="Comunidad", color="Tiempo", title="") +
  stat_ellipse(aes(color=Tiempo, group=Tiempo), level=0.95, linetype="solid", linewidth=1) +  
  geom_point() +  
  labs (
    shape = "Time",
     color = "Community")+
  scale_shape_manual(values = c(0:9)) + 
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink", "brown", "yellow", "cyan", "black")) +
  guides(color = guide_legend(nrow = 5), shape = guide_legend(nrow = 5)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 0, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(margin = margin(b = 10)))
  
ggsave("graficos/bray_tiempo.svg", bray_tiempo, width = 8, height = 6, dpi = 300)

bray_paneles <-  (bray_Comunidad | bray_tiempo) +
  plot_annotation(
    tag_levels = 'A',
    tag_prefix = "",
    tag_suffix = ") "
  )

ggsave("graficos/bray_paneles.svg", bray_paneles, width = 14, height = 6, dpi = 300)

```

### Microbial composition by Community - Bray-Curtis 

Here, microbial composition is represented using NMDS with Bray-Curtis distances, separating the data by Community type and grouping them by sampling time.

```{r}


Comunidades <- unique(sample_data(ps_prop)$Comunidad)
lista_graficos <- list()


for (s in Comunidades) {
  ps_Comunidad <- subset_samples(ps_prop, Comunidad == s)
  set.seed(123)
  ord_Comunidad <- ordinate(ps_Comunidad, method = "NMDS", distance = "bray")

  Bray<- plot_ordination(ps_Comunidad, ord_Comunidad, shape = "Tiempo", color = "Tiempo", title = paste("NMDS Bray-Curtis - Comunidad", s)) +
    stat_ellipse(aes(color = Tiempo, group = Tiempo), level = 0.95, linetype = "solid", linewidth = 1) +
    geom_point(size = 3) +  
  labs (
    shape = "Time",
     color = "Community")+
    scale_shape_manual(values = c(0:9)) +
    scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink", "brown", "yellow", "cyan", "black")) +
    guides(color = guide_legend(nrow = 5), shape = guide_legend(nrow = 5)) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      strip.text = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(margin = margin(b = 10)))
  
  file_nmds <- file.path(out_graf, paste0("Bray_comunidad_", s, ".svg"))
  ggsave(filename = file_nmds, plot =  Bray, width = 7, height = 5, dpi = 300)
 
  lista_graficos[[s]] <- Bray

  
  message("Guardado: ", file_nmds)
  
}

Bray_Comunidades_combinados <- wrap_plots(lista_graficos, ncol = 4) +
  plot_annotation(
    tag_levels = "A",
    tag_suffix = ") "
  )

ggsave(file.path(out_graf, "Bray_comunidades_combinadas.svg"),
       plot = Bray_Comunidades_combinados,
       width = 16, height = 12, dpi = 300)
```

### Redundancy analysis - Bray-Curtis 

A partial distance-based redundancy analysis (dbRDA) using Bray-Curtis distances was performed, modeling sample composition as a function of Time while controlling for Community, and as a function of Community while controlling for Time.

```{r}

Z <- as(otu_table(ps_prop), "matrix")
if (taxa_are_rows(ps_prop)) Z <- t(Z)

bray <- vegdist(Z, method = "bray")

meta <- data.frame(sample_data(ps_prop))
meta$Comunidad <- factor(meta$Comunidad)
meta$Tiempo <- as.numeric(meta$Tiempo)

mod_db_T <- capscale(bray ~ Tiempo + Condition(Comunidad), data = meta)
anova(mod_db_T)        
RsquareAdj(mod_db_T)

mod_db_C <- capscale(bray ~ Comunidad + Condition(Tiempo), data = meta)
anova(mod_db_C)        
RsquareAdj(mod_db_C)

```

## Microbial composition - UniFrac 

Here, microbial composition is represented using NMDS based on UniFrac distances, separating the data by Community type and sampling time, and visually grouping them according to these variables.

```{r}

set.seed(123)
ord_nmds_unifrac <- ordinate(new_ps, method = "NMDS", distance = "Unifrac", weighted=TRUE)      


Unifrac_Comunidad <- plot_ordination(new_ps, ord_nmds_unifrac, shape="Tiempo", color="Comunidad", title="") +
  stat_ellipse(aes(color=Comunidad, group=Comunidad), level=0.95, linetype="solid", linewidth=1) +  
  geom_point() +  
  labs (
    shape = "Time",
     color = "Community")+ 
  scale_shape_manual(values = c(0:9)) + 
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink", "brown", "yellow", "cyan", "black")) +
  guides(color = guide_legend(nrow = 5), shape = guide_legend(nrow = 5)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 0, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(margin = margin(b = 10)))


ggsave("graficos/Unifrac_Comunidad.svg", plot = Unifrac_Comunidad, width = 8, height = 6, dpi = 300)

Unifrac_Tiempo <- plot_ordination(new_ps, ord_nmds_unifrac, shape="Comunidad", color="Tiempo", title="") +
  stat_ellipse(aes(color=Tiempo, group=Tiempo), level=0.95, linetype="solid", linewidth=1) +  
  geom_point() +  
  labs (
    shape = "Time",
     color = "Community")+ 
  scale_shape_manual(values = c(0:9)) + 
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink", "brown", "yellow", "cyan", "black")) +
  guides(color = guide_legend(nrow = 5), shape = guide_legend(nrow = 5)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 0, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(margin = margin(b = 10)))


ggsave("graficos/Unifrac_Tiempo.svg", plot = Unifrac_Tiempo, width = 8, height = 6, dpi = 300)


Unifrac_paneles <-  (Unifrac_Comunidad |Unifrac_Tiempo) +
  plot_annotation(
    tag_levels = 'A',
    tag_prefix = "",
    tag_suffix = ") "
  )

ggsave("graficos/Unifrac_paneles.svg", Unifrac_paneles, width = 14, height = 6, dpi = 300)

```


### Microbial composition by Community - UniFrac 

Here, we represent microbial composition using NMDS based on UniFrac distances, separating the data by each Community type and grouping them according to sampling time.

```{r}

Comunidades <- unique(sample_data(ps_prop)$Comunidad)
lista_graficos <- list()


for (s in Comunidades) {
  ps_Comunidad_new <- subset_samples(new_ps, Comunidad == s)
  set.seed(123)
  ord_Comunidad_new <- ordinate(ps_Comunidad_new, method = "NMDS",distance = "Unifrac", weighted=TRUE) 

  Unifrac<- plot_ordination(ps_Comunidad_new, ord_Comunidad_new, shape = "Tiempo", color = "Tiempo", title = paste("NMDS Unifrac - Comunidad", s)) +
    stat_ellipse(aes(color = Tiempo, group = Tiempo), level = 0.95, linetype = "solid", linewidth = 1) +
    geom_point(size = 3) +  
  labs (
    shape = "Time",
     color = "Community") +
    scale_shape_manual(values = c(0:9)) +
    scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink", "brown", "yellow", "cyan", "black")) +
    guides(color = guide_legend(nrow = 5), shape = guide_legend(nrow = 5)) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      strip.text = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    legend.title = element_text(margin = margin(b = 10)))
  
  file_nmds <- file.path(out_graf, paste0("Unifrac_comunidad_", s, ".svg"))
  ggsave(filename = file_nmds, plot =  Unifrac, width = 7, height = 5, dpi = 300)
 
  lista_graficos[[s]] <- Unifrac

  
  message("Guardado: ", file_nmds)
  
}

Unifrac_Comunidades_combinados <- wrap_plots(lista_graficos, ncol = 4) +
  plot_annotation(
    tag_levels = "A",
    tag_suffix = ") "
  )

ggsave(file.path(out_graf, "Unifrac_comunidades_combinadas.svg"),
       plot = Unifrac_Comunidades_combinados,
       width = 16, height = 12, dpi = 300)

```
 
### Redundancy analysis - UniFrac

A partial distance-based redundancy analysis (partial dbRDA) using UniFrac distances was performed, modeling sample composition as a function of Time while controlling for Community, and as a function of Community while controlling for Time.

```{r}
Uni <- phyloseq::distance(new_ps, method = "unifrac", weighted = TRUE)

meta_Uni <- data.frame(sample_data(new_ps))
meta_Uni$Comunidad <- factor(meta_Uni$Comunidad)
meta_Uni$Tiempo <- as.numeric(meta_Uni$Tiempo)
mod_db_uni<- capscale(Uni ~ Tiempo + Condition(Comunidad), data = meta_Uni)
anova(mod_db_uni)             


mod_db_uni_C<- capscale(Uni ~ Comunidad + Condition(Tiempo), data = meta_Uni)
anova(mod_db_uni_C)             

```

## Analysis of the evolution of intra-Community dissimilarity over time using Bray-Curtis distances 

This section analyzes the temporal dynamics of microbial composition using the Bray-Curtis distance matrix. The mean distance within each Community is then calculated for each Time level, and the temporal trend is estimated for each Community. Next, the slopes are weighted to calculate a global statistic, and its statistical significance is evaluated using a restricted permutation test.

```{r}

bray_mat <- as.matrix(bray)


intra_bray <- sapply(seq_len(nrow(bray_mat)), function(i) {
  others <- which(meta$Comunidad == meta$Comunidad[i] &
                  meta$Tiempo    == meta$Tiempo[i] &
                  seq_len(nrow(bray_mat)) != i)
  if (length(others) < 1) return(NA_real_)   # n = 1 -> NA
  mean(bray_mat[i, others], na.rm = TRUE)
})

df_bray <-meta

df_bray$Tiempo <- factor(df_bray$Tiempo, levels = sort(unique(df_bray$Tiempo)))

Boxplot_bray<- ggplot(df_bray, aes(x = factor(Tiempo), y = intra_bray)) +
  geom_boxplot(outlier.shape = NA, coef = 1.5, fill = "lightblue") +
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkblue") +
  facet_wrap(~Comunidad, scales = "free_y") +
  labs(x = "Time", y = "Intra-community Bray-Curtis distance") +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "lightgray", colour = "black", size = 0.5),
    strip.text = element_text(margin = margin(t = 1, r = 2, b = 1, l = 2))
  )

print(Boxplot_bray)


ggsave("graficos/Boxplot_bray.svg",
       plot = Boxplot_bray,
       width = 10, height = 5, dpi = 300)

## -------------------------
## 1) Utilities
## -------------------------
to_matrix <- function(bray_mat){
  if (inherits(bray_mat, "dist")) {
    as.matrix(bray_mat)
  } else {
    as.matrix(bray_mat)
  }
}
 
## Mean distance within each (Community x time)
## number of pairs used (for weighting)
mean_within_time <- function(Dmat, meta){
  stopifnot(nrow(Dmat) == nrow(meta))
  # indices by community x time
  grp <- interaction(meta$Comunidad, meta$Tiempo, drop = TRUE)
  idx_list <- split(seq_len(nrow(Dmat)), grp)
 
  out <- lapply(names(idx_list), function(nm){
    ix <- idx_list[[nm]]
    # We need at least 2 replicates to form pairs
    if (length(ix) < 2) return(NULL)
    Di <- Dmat[ix, ix, drop = FALSE]
   # upper-triangle mean
    m <- mean(Di[upper.tri(Di)])
    n_pairs <- choose(length(ix), 2)
    parts <- strsplit(nm, split="\\.")[[1]]
    data.frame(Comunidad = parts[1],
               Tiempo = as.numeric(paste(parts[-1], collapse=".")), # soporta tiempos no enteros
               mean_dist = m,
               n_pairs = n_pairs,
               stringsAsFactors = FALSE)
  })
  do.call(rbind, out)
}
 
## Slope by community (WLS by number of pairs)
slope_by_Comunidad <- function(df_ct){
  split(df_ct, df_ct$Comunidad) |>
    lapply(function(dd){
      dd <- dd[order(dd$Tiempo), ]
     # if there are fewer than 2 time points with data, we cannot estimate a slope
      if (nrow(dd) < 2) return(data.frame(slope = NA_real_,
                                          n_pairs_total = sum(dd$n_pairs)))
      fit <- try(suppressWarnings(
        lm(mean_dist ~ Tiempo, data = dd, weights = n_pairs)
      ), silent = TRUE)
      if (inherits(fit, "try-error")) {
        data.frame(slope = NA_real_, n_pairs_total = sum(dd$n_pairs))
      } else {
        data.frame(slope = unname(coef(fit)["Tiempo"]),
                   n_pairs_total = sum(dd$n_pairs))
      }
    }) |>
    (\(lst) {
      ans <- do.call(rbind, lst)
      ans$Comunidad <- rownames(ans)
      rownames(ans) <- NULL
      ans
    })()
}
 
## Global statistic: weighted mean of slopes by total number of pairs
global_stat <- function(tab_slope){
  w <- tab_slope$n_pairs_total
  s <- tab_slope$slope
  # If there are NAs, ignore those communities (insufficient data)
  keep <- is.finite(s) & is.finite(w) & w > 0
  sum(s[keep] * w[keep]) / sum(w[keep])
}
 
## Pipeline to obtain the observed statistic
compute_observed <- function(Dmat, meta){
  df_ct <- mean_within_time(Dmat, meta)
  tab_slope <- slope_by_Comunidad(df_ct)
  list(df_ct = df_ct,
       tab_slope = tab_slope,
       stat_global = global_stat(tab_slope))
}
 
## Permutación: baraja 'time' **dentro de cada comunidad**
permute_times_within_comm <- function(meta){
  meta_perm <- meta
  meta_perm$Tiempo <- ave(meta$Tiempo, meta$Comunidad, FUN = function(t) sample(t))
  meta_perm
}
 
## -------------------------
## 2) Data preparation
## -------------------------
## Align meta and Dmat
Dmat <- to_matrix(bray_mat)
 
## Ensure that meta$Nombre aligns with the rows/columns of Dmat
## (we assume rownames(Dmat) are Names)
stopifnot(all(meta$Nombre %in% rownames(Dmat)))
## Reorder meta and D so that they match
meta <- meta[match(rownames(Dmat), meta$Nombre), ]
stopifnot(all(meta$Nombre == rownames(Dmat)))
 
## Coercions
meta$Comunidad <- as.factor(meta$Comunidad)
meta$Tiempo <- as.numeric(meta$Tiempo)
 
## -------------------------
## 3) Observed
## -------------------------
obs <- compute_observed(Dmat, meta)
obs$tab_slope        # slopes by community
obs$stat_global      # global statistic (we expect it to be negative if there is convergence)
 
## -------------------------
## 4) Restricted permutation (global p-value)
## -------------------------
set.seed(123)
B <- 4999  # number of permutations (adjust according to available time)
 
null_stats <- replicate(B, {
  meta_p <- permute_times_within_comm(meta)
  compute_observed(Dmat, meta_p)$stat_global
})
 
## p-valor unidireccional (H1: stat_global_obs < 0)
p_global <- (1 + sum(null_stats <= obs$stat_global)) / (B + 1)
 
## -------------------------
## 5) Results
## -------------------------
cat(sprintf("Global statistic (weighted mean slope): %.6f\n", obs$stat_global))
cat(sprintf("p-value (one-sided; convergence: slope < 0): %.4f\n", p_global))
 
## (Optional) Summary by community:
print(obs$tab_slope[order(obs$tab_slope$slope), ], row.names = FALSE)
 
## (Optional) Quick visualization of null vs observed
 hist(null_stats, breaks=40, main="Null distribution of the global statistic",
      xlab="Pendiente media ponderada")
 abline(v = obs$stat_global, col=2, lwd=2)
 
```

## Analysis of the evolution of intra-Community dissimilarity over time using UniFrac distances 

This section analyzes the temporal dynamics of microbial composition using the UniFrac distance matrix. The mean within-Community distance is calculated for each Time level, and the temporal trend is estimated for each Community. Next, the slopes are weighted to calculate a global statistic, and its statistical significance is evaluated using a restricted permutation test.

```{r}

Uni_mat <- as.matrix(Uni)

intra_Uni <- sapply(seq_len(nrow(Uni_mat)), function(i) {
  others <- which(meta_Uni$Comunidad == meta_Uni$Comunidad[i] &
                  meta_Uni$Tiempo    == meta_Uni$Tiempo[i] &
                  seq_len(nrow(Uni_mat)) != i)
  if (length(others) < 1) return(NA_real_)   # n = 1 -> NA
  mean(Uni_mat[i, others], na.rm = TRUE)
})

df_Uni <-meta_Uni
df_Uni$intra_Uni <- intra_Uni
df_Uni$Tiempo <- factor(df_Uni$Tiempo, levels = sort(unique(df_Uni$Tiempo)))

Boxplot_Uni<- ggplot(df_Uni, aes(x = factor(Tiempo), y = intra_Uni)) +
  geom_boxplot(outlier.shape = NA, coef = 1.5, fill = "lightblue") +
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkblue") +
  facet_wrap(~Comunidad, scales = "free_y") +
  labs(x = "Time", y = "Intra-community UniFrac distance") +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "lightgray", colour = "black", size = 0.5),
    strip.text = element_text(margin = margin(t = 1, r = 2, b = 1, l = 2))
  )

print(Boxplot_Uni)


ggsave("graficos/Boxplot_Uni.svg",
       plot = Boxplot_Uni,
       width = 10, height = 5, dpi = 300)

## -------------------------
## 1) Utilities
## -------------------------
to_matrix <- function(D){
  if (inherits(D, "dist")) {
    as.matrix(D)
  } else {
    as.matrix(D)
  }
}
 
## Mean distance within each (Community x time)
## number of pairs used (for weighting)
mean_within_time <- function(Dmat, meta){
  stopifnot(nrow(Dmat) == nrow(meta))
  # indices by community x time
  grp <- interaction(meta$Comunidad, meta$Tiempo, drop = TRUE)
  idx_list <- split(seq_len(nrow(Dmat)), grp)
 
  out <- lapply(names(idx_list), function(nm){
    ix <- idx_list[[nm]]
    # We need at least 2 replicates to form pairs
    if (length(ix) < 2) return(NULL)
    Di <- Dmat[ix, ix, drop = FALSE]
   # upper-triangle mean
    m <- mean(Di[upper.tri(Di)])
    n_pairs <- choose(length(ix), 2)
    parts <- strsplit(nm, split="\\.")[[1]]
    data.frame(Comunidad = parts[1],
               Tiempo = as.numeric(paste(parts[-1], collapse=".")), # soporta tiempos no enteros
               mean_dist = m,
               n_pairs = n_pairs,
               stringsAsFactors = FALSE)
  })
  do.call(rbind, out)
}
 
## Slope by community (WLS by number of pairs)
slope_by_Comunidad <- function(df_ct){
  split(df_ct, df_ct$Comunidad) |>
    lapply(function(dd){
      dd <- dd[order(dd$Tiempo), ]
      # if there are fewer than 2 time points with data, we cannot estimate a slope
      if (nrow(dd) < 2) return(data.frame(slope = NA_real_,
                                          n_pairs_total = sum(dd$n_pairs)))
      fit <- try(suppressWarnings(
        lm(mean_dist ~ Tiempo, data = dd, weights = n_pairs)
      ), silent = TRUE)
      if (inherits(fit, "try-error")) {
        data.frame(slope = NA_real_, n_pairs_total = sum(dd$n_pairs))
      } else {
        data.frame(slope = unname(coef(fit)["Tiempo"]),
                   n_pairs_total = sum(dd$n_pairs))
      }
    }) |>
    (\(lst) {
      ans <- do.call(rbind, lst)
      ans$Comunidad <- rownames(ans)
      rownames(ans) <- NULL
      ans
    })()
}
 
## Global statistic: weighted mean of slopes by total number of pairs
global_stat <- function(tab_slope){
  w <- tab_slope$n_pairs_total
  s <- tab_slope$slope
  # If there are NAs, ignore those communities (insufficient data)
  keep <- is.finite(s) & is.finite(w) & w > 0
  sum(s[keep] * w[keep]) / sum(w[keep])
}
 
## Pipeline to obtain the observed statistic
compute_observed <- function(Dmat, meta){
  df_ct <- mean_within_time(Dmat, meta)
  tab_slope <- slope_by_Comunidad(df_ct)
  list(df_ct = df_ct,
       tab_slope = tab_slope,
       stat_global = global_stat(tab_slope))
}
 
## Permutation: shuffle 'time' **within each community**
permute_times_within_comm <- function(meta){
  meta_perm <- meta
  meta_perm$Tiempo <- ave(meta$Tiempo, meta$Comunidad, FUN = function(t) sample(t))
  meta_perm
}
 
## -------------------------
### 2) Data preparation
## -------------------------
## Align meta and Dmat
Dmat_Uni <- to_matrix(Uni_mat)
 
## Ensure that meta$Nombre aligns with the rows/columns of Dmat
## (we assume rownames(Dmat) are Names)
stopifnot(all(meta_Uni$Nombre %in% rownames(Dmat_Uni)))
## Reorder meta and D so that they match
meta_Uni <- meta_Uni[match(rownames(Dmat_Uni), meta_Uni$Nombre), ]
stopifnot(all(meta_Uni$Nombre == rownames(Dmat_Uni)))
 
## Coercions
meta_Uni$Comunidad <- as.factor(meta_Uni$Comunidad)
meta_Uni$Tiempo <- as.numeric(meta_Uni$Tiempo)
 
## -------------------------
## 3) Observed
## -------------------------
obs <- compute_observed(Dmat_Uni, meta_Uni)
obs$tab_slope        # slopes by community
obs$stat_global      # global statistic (we expect it to be negative if there is convergence)
 
## -------------------------
## 4)  Restricted permutation (global p-value)
## -------------------------
set.seed(123)
B <- 4999  # number of permutations (adjust according to available time)
 
null_stats <- replicate(B, {
  meta_p <- permute_times_within_comm(meta_Uni)
  compute_observed(Dmat_Uni, meta_p)$stat_global
})
 
## One-sided p-value (H1: stat_global_obs < 0)
p_global <- (1 + sum(null_stats <= obs$stat_global)) / (B + 1)
 
## -------------------------
## 5) Results
## -------------------------
cat(sprintf("Global statistic (weighted mean slope): %.6f\n", obs$stat_global))
cat(sprintf("p-value (one-sided; convergence: slope < 0): %.4f\n", p_global))
 
## (Optional) Summary by community:
print(obs$tab_slope[order(obs$tab_slope$slope), ], row.names = FALSE)
 
## (Optional) Quick visualization of null vs observed
 hist(null_stats, breaks=40, main="Null distribution of the global statistic",
      xlab="Pendiente media ponderada")
 abline(v = obs$stat_global, col=2, lwd=2)
 
```

## Phylogenetic clustering analysis

The standardized effect size of the mean pairwise phylogenetic distance (MPD SES, from Mean Pairwise Distance Standardized Effect Size) between pairs of co-occurring taxa was determined. These values were then plotted for each community.


```{r}

set.seed(134)
# Create community table
comm <- as(otu_table(new_ps), "matrix")
if (taxa_are_rows(new_ps)) comm <- t(comm)

tree_pruned <- drop.tip(phy_tree(new_ps), setdiff(phy_tree(new_ps)$tip.label, colnames(comm)))
tree_dist <- cophenetic(tree_pruned)

mpd_res <- picante::ses.mpd(samp = comm, dis = tree_dist, null.model = "sample.pool", runs = 999)
saveRDS(mpd_res, file = "mpd_res_ps_new.rds")




# Add z-score to new_ps sample_data
sample_data(new_ps)$mpd.obs.z_rare <- mpd_res$mpd.obs.z
meta_new <- as(sample_data(new_ps), "data.frame")
meta_new$Comunidad <- factor(meta_new$Comunidad)
meta_new$mpd.obs.z_rare <- sample_data(new_ps)$mpd.obs.z_rare
meta_new$Tiempo  <- as.integer(meta_new$Tiempo) 
meta_new$cTiempo <- as.numeric(scale(meta_new$Tiempo, center = TRUE, scale = TRUE))

meta_new$Comunidad <- relevel(meta_new$Comunidad, ref = "A")
contrasts(meta_new$Comunidad)

lmm_mpd <- lm(mpd.obs.z_rare ~ cTiempo + Comunidad,  data = meta_new)
summary(lmm_mpd)
anova(lmm_mpd)


MPD_SES<- meta_new %>%
    ggplot(aes(x = as.factor(Tiempo), y = mpd.obs.z_rare, fill = Comunidad)) +
  geom_hline(yintercept = 0, lty = 'dashed') +
  geom_boxplot() +
  ylab('MPD SES') +
  geom_point(position = position_jitterdodge(), alpha = 0.3) +
  facet_wrap(~Comunidad, ncol = 3) +
  ggtitle('') +
  theme_bw() +
  labs(x = "Time",
       fill = "Community") +
    theme(
      axis.text.x = element_text(size = 12, hjust = 1),  
      axis.text.y = element_text(size = 12),
      axis.title.x = element_text(size = 12, face = "bold"),         
      axis.title.y = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14, face="bold"),
      plot.title   = element_text(size = 14, face = "bold", hjust = 0.5) 
    )

 MPD_SES
ggsave("graficos/MPD.png", MPD_SES, width = 16, height = 12, dpi = 300)

```


